QuickSched: Task-based parallelism with dependencies and conflicts

### 2013-10-28 paper/paper.tex
backup.

\documentclass[prodmode,acmtecs]{acmsmall} % Aptara syntax

@@

Conflicts can be modeled as exclusive locks on a shared resource
which have to be obtained by a task before it can execute.
Thus, in \fig{TaskConflicts}, before executing, task $F$ has
to obtain an exclusive lock on the resource associated with
the conflict between tasks $F$, $H$, and $I$.
While task $F$ is being executed, neither $H$ nor $I$ can 
lock the same resource, and therefore will not execute until
task $F$ is done and the lock has been released.

@@

This paper presents QuickSched, a framework for task-based
parallel programming with constraints.
Section~2 describes the underlying algorithms and data structures,
and Section~3 describes their specific implementation in QuickSched.
Section~4 presents two test-cases:
\begin{itemize}
    \item The tiled QR
    decomposition described in \cite{ref:Buttari2009} and for
    which the QUARK scheduler was originally developed,
    \item A task-based Barnes-Hut tree-code to compute the
    gravitational N-body problem,
\end{itemize}
These real-world examples show how QuickSched can be used in practice,
and can be used to assess its efficiency.
Section~5 concludes with some general observations and future work.

@@

\subsection{Tasks}

@@

The scheduler object is used as the main interface to the
QuickSched task scheduler, and as such contains the instances
other three object types:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct qsched {
    struct task *tasks;
    struct queue *queues;
    struct resource *res;
    int nr_tasks, nr_queues, nr_resources;
    volatile int waiting;
    };
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where\dots
Note that for brevity, and to avoid conflicts with the naming
schemes of other standard libraries, the type name {\tt qsched}
is used.

\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void qsched_start ( qsched *s ) {
    set the task waits and weights.
    s->waiting = s->nr_tasks;
    for ( k = 0 ; k < s->nr_tasks && s->tasks[tid[k]].wait == 0 ; k++ )
        qsched_enqueue( s , tid[k] );
    }
    \end{lstlisting}
\end{minipage}\end{center}

\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void qsched_enqueue ( qsched *s , struct task *t ) {
    int k, qid, score[ s->nr_queues ];
    for ( k = 0 ; k < s->nr_queues ; k++ )
        score[k] = 0;
    for ( j = 0 ; j < t->nr_locks ; j++ )
        score[ t->locks[j]->owner ] += 1;
    for ( j = 0 ; j < t->nr_uses ; j++ )
        score[ t->uses[j]->owner ] += 1;
    for ( qid = 0 , k = 1 ; k < s->nr_queues ; k++ )
        if ( score[k] > score[qid] )
            qid = k;
    queue_put( &s->queues[qid] , t );
    }
    \end{lstlisting}
\end{minipage}\end{center}

\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct task *qsched_gettask ( qsched *s , int qid ) {
    struct task *res;
    int k;
    while ( s->waiting ) {
        if ( ( res = queue_get( s->queues[qid] ) ) == NULL ) {
            k = id of queue with largest count.
            if ( ( res = queue_get( s->queues[k] ) ) != NULL )
                break;
            }
        else
            break;
        }
    if ( res != NULL ) {
        for ( k = 0 ; k < res->nr_locks ; k++ )
            res->locks[k]->owner = qid;
        for ( k = 0 ; k < res->nr_uses ; k++ )
            res->uses[k]->owner = qid;
        }
    return res;
    }
    \end{lstlisting}
\end{minipage}\end{center}

\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct task *qsched_done ( qsched *s , struct task *t ) {
    int k;
    for ( k = 0 ; k < t->nr_unlocks ; k++ )
        if ( atomic_dec( &t->unlocks[k]->wait ) == 1 )
            qsched_enqueue( s , t->unlocks[k] );
    atomic_dec( &s->waiting );
    }
    \end{lstlisting}
\end{minipage}\end{center}

\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct task *qsched_run ( qsched *s , void (*fun)( int , void * ) ) {
    #pragma omp parallel
    {
        int qid = omp_get_thread_num() % s->nr_queues;
        struct task *t;
        while ( ( t = qsched_gettask( s , qid ) ) != NULL ) {
            fun( t->type , t->data );
            qsched_done( s , t );
            }
        }
    }
    \end{lstlisting}
\end{minipage}\end{center}

@@

% Acknowledgments
\begin{acks}
The author would like to thank Sam Townsend for implementing the
kernels in the task-based QR decomposition as part of his MSc Thesis
at Durham University, as well as Lydia Heck of the Institute for
Computational Cosmology at Durham University for providing access
to, and expertise on, the COSMA cluster.
\end{acks}

>>>>

\documentclass[prodmode,acmtopc]{acmsmall} % Aptara syntax

@@

@@

Conflicts can be modeled directly as exclusive locks on a shared resource
which have to be obtained by a task before it can execute.
Thus, in \fig{TaskConflicts}, before executing, task $F$ has
to obtain an exclusive lock on the resource associated with
the conflict between tasks $F$, $H$, and $I$.
While task $F$ is being executed, neither $H$ nor $I$ can 
lock the same resource, and therefore will not execute until
task $F$ is done and the lock has been released.

This paper presents QuickSched, a framework for task-based
parallel programming with constraints.
In order to be of practical use, and to scale well with modern
multi-core shared-memory architectures, the task scheduler
must be:
\begin{itemize}
    \item {\em Correct}: All constraints, i.e.~dependencies and
        conflicts, must be correctly enforced,
    \item {\em Fast}: The overheads associated with task management
        should be as small as possible,
    \item {\em Memory/cache-efficient}: Tasks accessing similar
        sets of data should be preferentially executed on the
        same core to preserve memory/cache locality as far as possible, and
    \item {\em Parallel-efficient}: Tasks should be executed in an order
        that sufficient work is available for all computatoinal
        threads at all times.
\end{itemize}
\noindent This paper presents QuickSched, a framework for task-based
parallel programming with constraints, which aims to achieve
all of these goals.
Section~2 describes the underlying algorithms and data structures,
and Section~3 describes their specific implementation in QuickSched.
Section~4 presents two test-cases:
\begin{enumerate}
    \item The tiled QR
    decomposition described in \cite{ref:Buttari2009} and for
    which the QUARK scheduler was originally developed,
    \item A task-based Barnes-Hut tree-code to compute the
    gravitational N-body problem,
\end{enumerate}
These real-world examples show how QuickSched can be used in practice,
and can be used to assess its efficiency.
Section~5 concludes with some general observations and future work.

@@

\subsection{Tasks}
\label{sec:tasks}

@@

The scheduler object is used as the main interface to the
QuickSched task scheduler, and as such contains the instances
other three object types:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct qsched {
    struct task *tasks;
    struct queue *queues;
    struct resource *res;
    int nr_tasks, nr_queues, nr_resources;
    volatile int waiting;
    };
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where\dots the only additional field {\tt waiting} is
used to keep track of the number of tasks that have not been
executed.
Note that for brevity, and to avoid conflicts with the naming
schemes of other standard libraries, the type name {\tt qsched}
is used.

The tasks are executed as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void qsched_run ( qsched *s , void (*fun)( int , void * ) ) {
    qsched_start( s );
    #pragma omp parallel
    {
        int qid = omp_get_thread_num() % s->nr_queues;
        struct task *t;
        while ( ( t = qsched_gettask( s , qid ) ) != NULL ) {
            fun( t->type , t->data );
            qsched_done( s , t );
            }
        }
    }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where {\tt qsched\_start} initializes the tasks and
fills the queues (line~1).
For simplicity, OpenMP \cite{ref:Dagum1998}, which is available
for most compilers, is used to create a parallel section
in which the code between lines~4 and~12 is executed
concurrently.
The parallel section consists of a loop (lines~7--10) in
which a task is acquired via the function {\tt qsched\_gettask}
and its type and data are passed to a user-supplied execution
function.
Once the task has been executed, it is returned to the
scheduler via the function {\tt qsched\_done}.
The loop terminates when the scheduler runs out of tasks,
i.e.~when {\tt qstack\_gettask} returns {\tt NULL}, and
the function exits once all the threads have exited their
loops.

At the start of a parallel computation, {\tt qsched\_start}
identifies the tasks that have no dependencies and sends them
to queues:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void qsched_start ( qsched *s ) {
    set the task waits and weights.
    s->waiting = s->nr_tasks;
    for ( k = 0 ; k < s->nr_tasks && s->tasks[tid[k]].wait == 0 ; k++ )
        qsched_enqueue( s , &s->tasks[tid[k]] );
    }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the operations in line~2 are described in \sect{tasks}.

The function {\tt qsched\_enqueue} tries to identify the best
queue for a given task by looking at which queues last used
the resources used and locked by the task, e.g.:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void qsched_enqueue ( qsched *s , struct task *t ) {
    int k, qid, score[ s->nr_queues ];
    for ( k = 0 ; k < s->nr_queues ; k++ )
        score[k] = 0;
    for ( j = 0 ; j < t->nr_locks ; j++ )
        score[ t->locks[j]->owner ] += 1;
    for ( j = 0 ; j < t->nr_uses ; j++ )
        score[ t->uses[j]->owner ] += 1;
    for ( qid = 0 , k = 1 ; k < s->nr_queues ; k++ )
        if ( score[k] > score[qid] )
            qid = k;
    queue_put( &s->queues[qid] , t );
    }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the array {\tt score} keeps a count of the
task resources ``owned'', or last used, by each queue.
In lines~9--12 the queue with the highest such score is 
chosen on which the task is then put (line~13).

The function {\tt qsched\_gettask} fetches a task from
one of the queues:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct task *qsched_gettask ( qsched *s , int qid ) {
    struct task *res = NULL;
    int k;
    while ( s->waiting ) {
        if ( ( res = queue_get( s->queues[qid] ) ) == NULL ) {
            k = id of queue with largest count.
            if ( ( res = queue_get( s->queues[k] ) ) != NULL )
                break;
            }
        else
            break;
        }
    if ( res != NULL ) {
        for ( k = 0 ; k < res->nr_locks ; k++ )
            res->locks[k]->owner = qid;
        for ( k = 0 ; k < res->nr_uses ; k++ )
            res->uses[k]->owner = qid;
        }
    return res;
    }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the parameter {\tt qid} is the index of the
prefered queue.
If the queue is empty, or all of the tasks in that queue had
unresolved conflicts, the scheduler uses {\em work stealing}
\cite{ref:Blumofe1999}, i.e.~it looks for the queue with the
largest task count, and attempts to steal a task from it
(line~7).
If a task could be obtained from any queue (line~13),
the resources it locks and uses are marked as now being owned
by the prefered queue (lines~14--17).
Finally, the task, or {\tt NULL} if no task could be obtained,
is returned.

The final step in a task's lifecycle is, on completion,
to unlock the tasks which depend on it.
This is handled by the function {\tt qsched\_done}:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void qsched_done ( qsched *s , struct task *t ) {
    int k;
    for ( k = 0 ; k < t->nr_unlocks ; k++ )
        if ( atomic_dec( &t->unlocks[k]->wait ) == 1 )
            qsched_enqueue( s , t->unlocks[k] );
    atomic_dec( &s->waiting );
    }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent If any of the unlocked tasks' wait counters
goes to zero (line~4), then the unlocked task is ready to
run and is immediately dispatched via {\tt qsched\_enqueue},
as described earlier.
Once all the dependent tasks have been unlocked, the
{\tt waiting} counter is decremented (line~6).

@@

% Acknowledgments
\begin{acks}
The author would like to thank Sam Townsend for implementing the
kernels in the task-based QR decomposition as part of his MSc Thesis
at Durham University.
The author also thanks Lydia Heck of the Institute for
Computational Cosmology at Durham University for providing access
to, and expertise on, the COSMA cluster used in the performance
evaluation.
\end{acks}
